Results of two multi-chords stellar occultations by dwarf planet (1) Ceres

### 2017-08-16 Ceres/Old_versions/Felipe_04abr2015/Felipe_04abr2015/CeresV8.tex
mod 2017 ago16

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[T1]{fontenc}
\usepackage{prettyref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{color}
\usepackage[breaklinks=true]{hyperref}

@@

\title[Results of two stellar occultations by Ceres]{Results of two multi-chord stellar occultations by dwarf planet (1) Ceres}
\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1,}$\thanks{E-mail: altair08@astro.ufrj.br},
B. L. Giacchini$^{2,3,4}$,%\thanks{E-mail: breno@cbpf.br},
F. Braga-Ribas$^{5, 6, \dag}$
M. Assafin$^{1, \dag, \ddag}$,\newauthor
R. Vieira-Martins$^{1,5, \dag, \ddag}$,
J.I.B. Camargo$^{5, \dag}$,
B. Sicardy$^{7}$,
B. Timerson$^{4}$, 
T. George$^{4}$, \newauthor
J. Broughton$^{8}$, 
T. Blank$^{4}$,
G. Benedetti-Rossi$^{5}$, 
J. Brooks$^{4}$,  
R. F. Dantowitz$^{9}$,  \newauthor
D. W. Dunham$^{4}$, 
J. B. Dunham$^{4}$, 
C. K. Ellington$^{4}$,
M. Emilio$^{10}$,
F.R. Herpich$^{11}$,  \newauthor
C. Jacques$^{12}$,
P. D. Maley$^{4,13}$,
L. Mehret$^{10}$,
A.J.T. Mello$^{11}$,
A.C. Milone$^{14}$,\newauthor
E. Pimentel$^{12}$,  
W. Schoenell$^{11}$,
N. S. Weber$^{9}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{7}$LESIA, Observatoire de Paris, CNRS UMR 8109, Universit\'{e} Pierre et Marie Curie, Universit\'{e} Paris-Diderot, 5 place Jules Janssen,\\ F-92195 Meudon Cedex, France\\
$^{8}$RASNZ Occultation Section, P.O. Box 3181, Wellington, New Zealand\\
$^{9}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
$^{10}$Universidade Estadual de Ponta Grossa, Ponta Grossa, Brazil\\
$^{11}$Universidade Federal de Santa Catarina, Florian\'opolis, Brazil\\
$^{12}$Centro de Estudos Astron\^omicos de Minas Gerais, Belo Horizonte, Brazil\\
$^{13}$NASA Johnson Space Center Astronomical Society, Houston, Texas, USA\\
$^{14}$Instituto Nacional de Pesquisas Espaciais, S\~ao Jos\'e dos Campos, Brazil\\
$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil\\
$^\ddag$Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France
}
\begin{document}

@@

\begin{abstract}
We report the results of two multi-chord stellar occultations by the dwarf planet (1) Ceres that were observed  from Brazil on 2010 August 17, and from the USA on 2013 October 25. Four positive detections were obtained for the 2010 occultation, and nine for the 2013 occultation. Elliptical models were adjusted to the observed chords to obtain Ceres' size and shape. Two limb fitting solutions were studied for each event. The first one is a nominal solution with an indeterminate polar aspect angle. The second one was constrained by the pole coordinates as given by \cite{Drummond2014}. Assuming a Maclaurin spheroid, we determine an equatorial diameter of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$ as our best solution. These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the NASA's \textit{Dawn} space mission.
\end{abstract}

@@

Ceres is the sole example of a dwarf planet in the inner Solar System. Far from being mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System. Indeed, it was proposed that Ceres' origin could have been as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}. Even if it was formed close to its current location, the dynamical history of the Solar System must have let its signatures on Ceres. These could include not only the late heavy bombardment features that might exist on its surface, but also the makeup of its volatiles, which could have been transported from the outer region.

Since the 1970's it has been speculated that Ceres could contain water ice, which was recently verified \citep{Kuppers2014}. Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer \citep{CastilloR2011}. Yet the very question of whether Ceres underwent differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, what its geological evolution was, and if it is still active. Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, which is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.

Containing approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e. a Maclaurin or a Jacobi ellipsoid. In fact, direct observations of Ceres by means of adaptive optics indicate that it is an oblate spheroid \citep{Drummond2014}. The precise knowledge of its size and shape is of utmost importance, for the models of density, internal structure and differentiation.% can depend on the size and oblateness of the body.

@@

The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of some kilometres, at a time when the uncertainties were often ten times larger. The high apparent brightness of Ceres, as compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted. For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed \citep{Dunham2014}. Two of them had only two chords each, thus not sufficient for providing accurate results\footnote{These events took place on 1994 August 22 and 2010 October 30.}. The two remaining events, which occurred on 2010 August 17 and 2013 October 25, are reported in the present work and are the first ones that used charged-couple devices (CCD) as recording systems. % The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords.
Throughout the paper we shall refer to these events as the `2010 occultation' and the `2013 occultation'.

@@

This work is organized as follows. In Sections 2 and 3 we analyse the 2010 and the 2013 occultations, respectively. In Section 4 we give the geocentric positions of Ceres derived from the occultations. The comparison of both results to those in the literature is carried out in Section 5.

@@

As predicted, on 2010 August 17 Ceres occulted the star TYC 6833-163-1 (UCAC4 313-111823), which has magnitude $V = 11.55$ and has the ICRS position, based on UCAC4 catalogue \citep{Zacharias2013}, to the date of occultation, corrected of proper motion:

%abaixo (UCAC2 20678706),
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0085\\
    \delta = -27\degr 26\arcmin 38\farcs 867.
  \end{array}
\right.
\label{Eq: 2010-coord}
\end{equation}

@@

\begin{figure}
\includegraphics[scale=0.42]{figures/Ceres_2010.eps} 
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event. The big red dot is the geocentric closest approach at 22:40:25 UT. The small red ones represent the centre of the shadow separated by one minute, shadow moves from the left to the right. Blue dots are the sites that have observed the event. As described in text, UFSC had a negative chord.
\label{Fig: Ceres-2010-map}}
\end{figure}

Observations were carried out in Brazil from five different sites as displayed in Table \ref{Tab: obs-2010} and Fig. \ref{Fig: Ceres-2010-map}. The occultation was detected from four of them. The southern-most one (UFSC) had a negative chord. From the positive ones, the one named INPE started observing after the event was already in progress, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.

A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km~s^{-1}$ in the plane of the sky. Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.

All the observations were made through a sequence of images obtained with CCDs. The times of each exposure were available on the header of each image with an internal accuracy of a few hundredths of a second. CEAMIG had only the integer part of the second available, due to the acquisition software used. The fraction of a second could be retrieved as described on the Section~3.1 of \cite{BragaRibas2013}  (which shall be referred as BR13 hereafter). Cycle times (exposure plus read-out) varied from 2 to 52 seconds, as can be verified on Table~\ref{Tab: obs-2010}, making it a heterogeneous set of observations and imposing an error of a few seconds for the ingress / egress times of some sites.

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope & Exposure & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & f-ratio & Camera & Egress    & \\          
\hline
 Belo Horizonte & 43$\degr$59\arcmin 51.1\arcsec~W & LX200 & 5 s & Positive & C. Jacques  \\
 CEAMIG-REA &19$\degr$49\arcmin 49.0\arcsec~S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 s &  E. Pimentel \\
            & 825 m       &  f/10         &   SBIG ST10        & 22:40:20 $\pm$ 5 s &   \\
 & & & & & \\
 Pico dos Dias    & 45$\degr$34\arcmin45.1\arcsec~W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
 LNA    &22$\degr$32\arcmin03.7\arcsec~S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 s &  G. B. Rossi \\
            & 1864 m     & f/12.5          &    Andor Ikon       & 22:41:55.3 $\pm$ 0.7 s &              \\
 & & & & & \\
 S\~ao Jos\'e dos       & 45$\degr$51\arcmin43.0\arcsec~W &  C11  & 2 s & Egress only & A. C. Milone\\
 Campos       &23$\degr$12\arcmin32.0\arcsec~S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
 INPE      & 620 m      & f/6.3          & SBIG ST7     & 22:42:03.0 $\pm$ 0.2 s & M. Okada    \\
 & & & & & \\
 Ponta Grossa       & 50$\degr$05\arcmin56.0\arcsec~W & RCX 400 &30 s & Positive & M. Emilio   \\
 UEPG       &25$\degr$05\arcmin22.2\arcsec~S & 40 cm & 52 s & 22:37:17 $\pm$ 13 s & L. Mehret   \\
            & 910 m      & f/8          & SBIG STL6E     & 22:39:56 $\pm$ 13 s & \\
 & & & & & \\
 Florian\'opolis       & 48$\degr$31\arcmin20.5\arcsec~W &   C11      & 3 s & No occultation  & W. Schoenell\\
 UFSC       &27$\degr$36\arcmin12.3\arcsec~S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
            & 20 m       & f/10          &  SBIG ST7      &   End obs.: 22:51:21      & F. R. Herpich \\
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

The flux of the star in the five occultation chords was obtained from the FITS images with the \textsc{platform for reduction of astronomical images automatically (praia)} \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of the star plus Ceres, as they were merged right before and after the occultation. Additionally,
%, with exception of INPE
 they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.

The ingress (disappearance) and egress (reappearance) instants of the occultation were obtained for each light curve by fitting a square-well model convoluted with the Fresnel diffraction, the CCD bandwidth, the stellar apparent diameter, and the applied finite exposure time; see \cite{Widemann2009} and BR13. %\cite{BragaRibas2013}.
%The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres' geocentric distance $D = 2.29 \text{ au} = 3.42 \times 10^{8}$~km is 0.3~km for a typical wavelength of $\lambda = 0.65$~$\micron$. The star diameter is estimated using the formulae of \cite{vanBelle1999}. Its $B$, $V$, and $K$ apparent magnitudes are 12.00, 11.78 and 9.18, respectively, in the Tycho catalog \citep{Hog2000}. This yields a diameter of about 0.11~km projected at Ceres' distance. 
The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9~km in the celestial plane. Therefore, the time error of ingress and egress are largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are on the order a few hundreds meters for this event.
The occultation data fit consists in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress or egress instant, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
%The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.

\begin{figure}
\includegraphics[scale=0.58]{figures/Ceres_2010_fluxratio.eps} 
\caption{The four occultation light curves normalized and vertically shifted by a factor of 0.2 for better visualisation. The solid black lines are the best fit of the square-well model to the data. Red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time. The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites. Exposures at INPE started after the immersion, as explained in the text. \label{Fig: Ceres-2010-curves}}
\end{figure}

@@

The methodology used to analyse Ceres' profile from the observations is the same described by \cite{Sicardy2011} and BR13.%\cite{BragaRibas2013}. 
 ~Each combination of site position and recorded ingress/egress instant, together with star coordinates and Ceres' ephemeris, corresponds to a point in the plane of the sky. The collection of all these points ideally determines the apparent limb of Ceres.
%~To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point $(f_{i,obs},g_{i,obs})$ on the plane of the sky. The collection of those points ideally determines the apparent limb of Ceres.

We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This choice is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star. 
Of course, the apparent semimajor axis {$a^\prime$} is equivalent to the equatorial radius $R_{equa}$ of the ellipsoid.
The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris \citep{Giorgini1996} and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east. The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. The best-fitting solution is obtained minimizing a reduced $\chi^{2}$ function, where we define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.  All the procedures that allow the determination of the error bars of the physical parameters can be found on BR13.

@@

Two possible solutions were considered for the limb fitting. The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts. As we shortly show, it led to a rather large uncertainty on the position angle. Furthermore, the nominal solution alone is not capable of returning the true oblateness, which can be evaluated through equation 2 of BR13  provided that the aspect angle $\zeta$ is known.% or constrained by some physical hypothesis.

>>>>

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[T1]{fontenc}
\usepackage{prettyref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{color}
\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line

@@

\title[Results of two stellar occultations by Ceres]{Results of two multi-chords stellar occultations by dwarf planet (1) Ceres}
\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1,}$\thanks{E-mail: altair08@astro.ufrj.br},
B. L. Giacchini$^{2,3,4}$,%\thanks{E-mail: breno@cbpf.br},
F. Braga-Ribas$^{5, 6, \dag}$
M. Assafin$^{1, \dag}$,\newauthor
R. Vieira-Martins$^{1,5, \dag}$,
J.I.B. Camargo$^{5, \dag}$,
B. Sicardy$^{7}$,
B. Timerson$^{4}$, 
T. George$^{4}$, \newauthor
J. Broughton$^{8}$, 
T. Blank$^{4}$,
G. Benedetti-Rossi$^{5}$, 
J. Brooks$^{4}$,  
R. F. Dantowitz$^{9}$,  \newauthor
D. W. Dunham$^{4}$, 
J. B. Dunham$^{4}$, 
C. K. Ellington$^{4}$,
M. Emilio$^{10}$,
F.R. Herpich$^{11}$,  \newauthor
C. Jacques$^{12}$,
P. D. Maley$^{4,13}$,
L. Mehret$^{10}$,
A.J.T. Mello$^{11}$,
A.C. Milone$^{14}$,\newauthor
E. Pimentel$^{12}$,  
W. Schoenell$^{11}$,
N. S. Weber$^{9}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{7}$LESIA, Observatoire de Paris, CNRS UMR 8109, Universit\'{e} Pierre et Marie Curie, Universit\'{e} Paris-Diderot, 5 place Jules Janssen,\\ F-92195 Meudon Cedex, France\\
$^{8}$RASNZ Occultation Section, 18 Branch Cr. Reedy Creek, QLD 4227, Australia\\
$^{9}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
$^{10}$Universidade Estadual de Ponta Grossa, Ponta Grossa, Brazil\\
$^{11}$Universidade Federal de Santa Catarina, Florian\'opolis, Brazil\\
$^{12}$Centro de Estudos Astron\^omicos de Minas Gerais, Belo Horizonte, Brazil\\
$^{13}$NASA Johnson Space Center Astronomical Society, 4535 Cedar Ridge Trail, Houston, TX 77059 USA\\
$^{14}$Instituto Nacional de Pesquisas Espaciais, São José dos Campos, Brazil\\
$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil
}
\begin{document}

@@

\begin{abstract}
We report the results of two multi-chord stellar occultations by the dwarf planet (1) Ceres that were observed  from Brazil on August 17 2010, and from USA on October 25 2013. Four positive detections were obtained for the 2010 occultation, and nine for the 2013 occultation. Elliptical models were adjusted to the observed chords to obtain the Ceres' size and shape. Two limb fitting solution were studied for each event. The first one is a nominal solution with an indeterminate polar aspect angle. The second one was constrained by the pole coordinates as given by \cite{Drummond2014}. Assuming a Maclaurin spheroid, we determine an equatorial diameter of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$ as our best solution. These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the Nasa's \textit{Dawn} space mission.
\end{abstract}

@@

Ceres is the sole exemplar of a dwarf planet in the inner Solar System. Far from being a mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System. Indeed, it was proposed that Ceres could have its origin as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}. It may well had been formed close to its current location, however, even on this scenario, the dynamical history of the Solar System must have let its signatures on Ceres. Not only on the late heavy bombardment features that might exist on its surface, but also on its volatiles, which could have been transported from the outer region.

Since the 1970's it has been speculated that Ceres could contain water ice, what was recently verified \citep{Kuppers2014}. Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer \citep{CastilloR2011}. Yet the very question of whether Ceres suffered differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, how its geological evolution was, and if it is still active. Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, that is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.

Owning approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e. a Maclaurin or a Jacobi ellipsoid. In fact, direct observations of Ceres by means of adaptive optics indicate that it is an oblate spheroid \citep{Drummond2014}. The precise knowledge of its size and shape is of utmost importance, for the models of density, internal structure and differentiation.% can depend on the size and oblateness of the body.

@@

The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of some kilometres, in a time when the uncertainties were often ten times larger. The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted. For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed. Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}. The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work and are the first ones that used charged-couple devices (CCD) as recording systems. % The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords.
Throughout the paper we shall refer to these events as the `2010 occultation' and the `2013 occultation'.

@@

This work is organized as follows. In Sections 2 and 3 we analyse, the 2010 and the 2013 occultations, respectively. In Section 4 we give the geocentric positions of Ceres derived from the occultations. The comparison of both results to those on the literature is carried out in Section 5.

@@

On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1, %(UCAC2 20678706),
 which has magnitude $V = 11.55$ and has the ICRS position:
% based on UCAC4 catalogue \citep{Zacharias2013} to the date of occultation, corrected of proper motion:

%abaixo (UCAC2 20678706),
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
    \delta = -27\degr 26\arcmin 38\arcsec.890.
  \end{array}
\right.
\label{Eq: 2010-coord}
\end{equation}

@@

\begin{figure}
\includegraphics[scale=0.42]{figures/Ceres_2010.png} 
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event. The big red dot is the geocentric closest approach at 22:40:25 UT. The small red ones represent the centre of the shadow separated by one minute, shadow moves from the left to the right. Blue dots are the sites that have observed the event. As described in text, UFSC had a negative chord.
\label{Fig: Ceres-2010-map}}
\end{figure}

Observations were carried out in Brazil from five different sites as displayed in Table \ref{Tab: obs-2010} and Fig. \ref{Fig: Ceres-2010-map}. The occultation was detected on four of them. The south most one (UFSC) had a negative chord. From the positive ones, the one named INPE started the observation when the event was already in progress, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.

A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km.s^{-1}$ in the plane of the sky. Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.

All the observations were made through a sequence of images obtained with CCDs. The times of each exposure were available on the header of each image with an internal accuracy of a few hundredths of a second. CEAMIG had only the integer part of the second available, due to the acquisition software used. The fraction of a second could be retrieved as described on the Section~3.1 of \cite{BragaRibas2013}  (which shall be referred as BR13 hereafter). Cycle times (exposure plus read-out) varied from 2 to 52 seconds, as can be verified on Table~\ref{Tab: obs-2010}, making of it an heterogeneous set of observations and imposing an error of a few seconds for the ingress / egress times of some sites.

@@

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope & Exposure & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & f-ratio & Camera & Egress    & \\          
\hline
 Belo Horizonte & 43$\degr$59'51.1'' W & LX200 & 5 s & Positive & C. Jacques  \\
 CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
            & 825 m       &  f/10         &   SBIG ST10        & 22:40:20 $\pm$ 5 &   \\
 & & & & & \\
 Pico dos Dias    & 45$\degr$34'45.1'' W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
 OPD/LNA    &22$\degr$32'03.7'' S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
            & 1864 m     & f/12.5          &    Andor Ikon       & 22:41:55.3 $\pm$ 0.7 &              \\
 & & & & & \\
 S\~ao Jos\'e dos       & 45$\degr$51'44.0'' W &  C11  & 2 s & Egress only & A. C. Milone\\
 Campos       &23$\degr$12'33.0'' S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
 INPE      & 975 m      & f/6.3          & SBIG ST7     & 22:42:03.0 $\pm$ 0.2 & M. Okada    \\
 & & & & & \\
 Ponta Grossa       & 50$\degr$05'56.0'' W & RCX 400 &30 s & Positive & M. Emilio   \\
 UEPG       &25$\degr$05'22.2'' S & 40 cm & 52 s & 22:37:17 $\pm$ 13 & L. Mehret   \\
            & 910 m      & f/8          & SBIG STL6E     & 22:39:56 $\pm$ 13 & \\
 & & & & & \\
 Florian\'opolis       & 48$\degr$31'20.5'' W &   C11      & 3 s & No occultation  & W. Schoenell\\
 UFSC       &27$\degr$36'12.3'' S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
            & 20 m       & f/10          &  SBIG ST7      &   End obs.: 22:51:21      & F. R. Herpich \\
\hline
\end{tabular}
\end{minipage}
\end{table*}

@@

The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of the star plus Ceres, as they were merged right before and after the occultation. Additionally,
%, with exception of INPE
 they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.

The start and end times of the occultation were obtained for each light curve by fitting a square-well model convoluted with the Fresnel diffraction, the CCD bandwidth, the stellar apparent diameter, and the applied finite exposure time; see \cite{Widemann2009} and BR13. %\cite{BragaRibas2013}.
%The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres' geocentric distance $D = 2.29 \text{ au} = 3.42 \times 10^{8}$~km is 0.3~km for a typical wavelength of $\lambda = 0.65$~$\micron$. The star diameter is estimated using the formulae of \cite{vanBelle1999}. Its $B$, $V$, and $K$ apparent magnitudes are 12.00, 11.78 and 9.18, respectively, in the Tycho catalog \citep{Hog2000}. This yields a diameter of about 0.11~km projected at Ceres' distance. 
The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9~km in the celestial plane. Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are on the order a few hundreds meters for this event.
The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
%The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.

\begin{figure}
\includegraphics[scale=0.58]{figures/Ceres_2010_fluxratio.png} 
\caption{The four occultation light curves normalized and vertically shifted by a factor of 0.2 for better visualisation. The solid black lines are the best fit of the square-well model to the data. Red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time. The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites. Exposures at INPE started after the immersion, as explained in the text. \label{Fig: Ceres-2010-curves}}
\end{figure}

@@

The methodology used to analyse Ceres' profile from the observations is the same described by \cite{Sicardy2011} and BR13.%\cite{BragaRibas2013}. 
 ~To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point on the plane of the sky. The collection of those points ideally determines the apparent limb of Ceres.
%~To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point $(f_{i,obs},g_{i,obs})$ on the plane of the sky. The collection of those points ideally determines the apparent limb of Ceres.

We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star. 
Of course, the apparent semimajor axis {$a^\prime$} is equivalent to the equatorial radius $R_{equa}$ of the ellipsoid.
The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#26 Ceres' ephemeris and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east. The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.  All the procedure that allows the determination of the error bars of the physical parameters can be found on BR13.

@@

Two possible solutions were considered for the limb fitting. The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts. As we shortly show, it led to a rather large uncertainty on the position angle. Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation 2 of BR13.% \ref{Eq: TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.
